Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-18 docs/manuscript.Rmd
updated

fig_nums(
	"steiger_sensitivity_plot", 
	caption="a) We can predict the values the Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface. When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by range of values where the blue surface is above the red plane, those values of measurement error lead to our observed Steiger test inferring the wrong causal direction. Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y. A measure of reliability, therefore, is the ratio of the negative and positive volumes of the total space bound by the blue and red surfaces, $R = \\frac{V_{z \\geq 0}}{ - V_{z < 0} }$. In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the $R = 4.40$, which means that 4.40 times as much of the possible measurement error values are in support of the $x \\rightarrow y$ direction of causality than $y \\rightarrow x$. b) The influence of $\\rho_{x,y}$ (x-axis) on the Steiger reliability values (y-axis) for different values of $\\rho_{g,x}^2$. Inferring the direction of causality is more sensitive to measurement error when there are stronger causal associations between X and Y.",
	display=FALSE
)

@@

@@

An established study design that can provide information about causality is randomisation. Given the hypothesis that trait A (henceforth referred to as the exposure) is causally related to trait B (henceforth referred to as the outcome), randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control. The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship. This provides the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure. The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2003; @DaveySmithHemani2014]. Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation-based approaches and Mendelian randomisation (MR).

@@

MR has some important advantages over the mediation-based approaches. The mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016]. Additionally, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method. We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a method to evaluate the sensitivity of the result of this extension to measurement error. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially. Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels. Our analyses highlight that because these different causal inference techniques have varying strengths and weaknesses, triangulation of evidence from as many sources as possible should be practiced in causal inference [@Lawlor2017].

@@

where $\beta_{MR} \neq 0$ denotes the existence of causality, and the $\beta_{MR}$ is an estimate of the causal effect.

@@

where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$. The same model of measurement error can be applied to the outcome variable $y$. In this study we assume there is no measurement error in the SNP.

@@

The method that we will now describe is able to distinguish between two models, $x \rightarrow y$ or $y \rightarrow	x$. Unlike the CIT framework, this approach cannot infer if the true model is $x \leftarrow g \rightarrow y$. We also assume all genetic effects are additive.

@@

The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$. However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} > \rho_{g,y_o}$. As we show in the appendix, the maximum measurement error of $x_o$ is $\rho_{g,x_o}$ and the minimum is 0 (same applies to $y_o$). It is possible to simulate what the inferred causal direction would be for all values within these bounds. To evaluate how reliable, $R$, the inference of the causal direction is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test. We also do the same for the measurement error values which are not compatible. The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction. A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality. Full details are provided in Appendix 2 and `r cfig("steiger_sensitivity_plot")`a.

@@

In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$. The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$. When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that

@@

If we do not know whether the SNP $g$ has a primary influence on $x$ or $y$ then CIT can attempt to infer the causal direction. Here we introduce the MR Steiger approach to also orient the direction of causality in an MR analysis when the underlying biology of the SNP is not fully understood.

We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2), where $d$ is a metric of the degree to which the Steiger test is liable to provide the wrong estimate (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction). Thus it is important to note that under certain conditions when measurement imprecision is present the Steiger test could make erroneous inference about the causal direction because the value of $d$ is not restricted to being greater than 0. However `r cfig("d_relationship_figure")` shows that when there is no measurement error in $x$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$. It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the Steiger test to measurement error is relatively low.

We performed simulations to explore the performance of this approach in comparison to CIT in terms of the rate at which a strong causal relationship is obtained for the correct direction of causality, and the rate at which a strong causal relationship is obtained where the reported direction of causality is incorrect. Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.

`r cfig("cit_mr_comparison_figure")`a shows that, for the "causal model", when $d < 0$ the MR analysis is indeed liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size. However, the CIT is in general more fallable to reporting a robust causal association for the wrong direction of causality. When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality. The CIT, unlike the Steiger test for MR, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to evaluate this confounding model as being causal.

@@

We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$). We then went on to predict the causal directions of the associations for varying levels of possible measurement error. `r cfig("shakhplot")`a shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.

>>>>

@@

fig_nums(
	"steiger_sensitivity_plot", 
	caption="a) We can predict the values the Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface. When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by range of values where the blue surface is above the red plane, those values of measurement error lead to our observed Steiger test inferring the wrong causal direction. Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y. A measure of reliability, therefore, is the ratio of the negative and positive volumes of the total space bound by the blue and red surfaces, $R = \\frac{V_{z \\geq 0}}{ - V_{z < 0} }$. In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the $R = 4.40$, which means that 4.40 times as much of the possible measurement error values are in support of the $x \\rightarrow y$ direction of causality than $y \\rightarrow x$. b) Plots depicting the parameter space in which the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$ is negative. When $d$ is negative the Steiger test is liable to infer the wrong direction of causality. Shaded regions show the parameter space where $d$ is negative. The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.
	",
	display=FALSE
)

@@

An established study design that can provide information about causality is randomisation. Given the hypothesis that trait A (henceforth referred to as the exposure) is causally related to trait B (henceforth referred to as the outcome), randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and treating the other as a control. The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship. This provides the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure. The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2003; @DaveySmithHemani2014]. Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation-based approaches and Mendelian randomisation (MR).

@@

MR has some important advantages over the mediation-based approaches. The mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016]. Additionally, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable [@Pierce2012]. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method. We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a matric to evaluate the sensitivity of the result of this extension to measurement error. Together these extensions improve the utility of MR in cases where mediation based methods might have otherwise been used preferentially. Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels. Our analyses highlight that because these different causal inference techniques have varying strengths and weaknesses, triangulation of evidence from as many sources as possible should be practiced in causal inference [@Lawlor2017].

@@

where $\beta_{MR} \neq 0$ denotes the existence of causality, and $\beta_{MR}$ is an estimate of the causal effect.

@@

where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$. The same model of measurement error can be applied to the outcome variable $y$. In this study we assume there is no measurement error in the SNP and measurement error in the exposure and the outcome are uncorrelated.

@@

The method that we will now describe is designed to distinguish between two models, $x \rightarrow y$ or $y \rightarrow	x$. Unlike the CIT framework, this approach cannot infer if the true model is $x \leftarrow g \rightarrow y$. We also assume all genetic effects are additive.

@@

The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$. However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} > \rho_{g,y_o}$. As we show in the appendix, the maximum measurement error of $x_o$ is $\rho_{g,x_o}$ and the minimum is 0 (same applies to $y_o$). It is possible to simulate what the inferred causal direction would be for all values within these bounds. To evaluate how reliable, $R$, the inference of the causal direction is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test. We also do the same for the measurement error values which are not compatible. The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction. A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality. Full details are provided in Appendix 2.

@@

In the causal inference test (CIT), the 4th condition (see Methods) employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$. When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that

@@

If we do not know whether the SNP $g$ has a primary influence on $x$ or $y$ then CIT can attempt to infer the causal direction. Here we introduce the MR Steiger approach to similarly orient the direction of causality but in an MR analysis when the underlying biology of the SNP is not fully understood.

For a particular association, it is of interest to identify the range of possible measurement error values agree and disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, Appendix 2). This metric can be used to evaluate the reliability of MR Steiger.

We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2) determines the range of parameters around which the Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction). `r cfig("steiger_sensitivity_plot")`b shows that when there is no measurement error in $x$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$. It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the Steiger test to measurement error is relatively low.

### Comparison of CIT and MR Steiger for obtaining the correct direction of causality

We used simulations to explore the performance of the MR Steiger approach in comparison to CIT in terms of the rate at which a strong causal relationship is obtained for the correct direction of causality, and the rate at which a strong causal relationship is obtained where the reported direction of causality is incorrect. Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.

`r cfig("cit_mr_comparison_figure")`a shows that, for the "causal model", when $d < 0$ the MR analysis is indeed liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size. However, the CIT is in general more fallable to reporting a robust causal association for the wrong direction of causality. When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality. The CIT, unlike the Steiger test for MR, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to evaluate erroneously this confounding model as being causal.

@@

We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$). We then went on to predict the causal directions of the associations for varying levels of possible measurement error. `r cfig("shakhplot")`a shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.
